Control of ciliary transcriptional programs during spermatogenesis by antagonistic transcription factors

Version 1

Abstract

Abstract

Existence of cilia in the last eukaryotic common ancestor (LECA) raises a fundamental question in biology: how the transcriptional regulation of ciliogenesis has evolved? One conceptual answer to this question is by an ancient transcription factor regulating ciliary gene expression in both unicellular and multicellular organisms, but examples of such transcription factors in eukaryotes are lacking. Previously, we showed that an ancient transcription factor XAP5 is required for flagellar assembly in Chlamydomonas. Here, we show that XAP5 and XAP5L are two conserved pairs of antagonistic transcription regulators that control ciliary transcriptional programs during spermatogenesis. Male mice lacking either XAP5 or XAP5L display infertility, as a result of meiotic prophase arrest and sperm flagella malformation, respectively. Mechanistically, XAP5 positively regulates the ciliary gene expression by activating the key regulators including FOXJ1 and RFX families during the early stage of spermatogenesis. In contrast, XAP5L negatively regulates the expression of ciliary genes via repressing these ciliary transcription factors during the spermiogenesis stage. Our results provide new insights into the mechanisms by which temporal and spatial transcription regulators are coordinated to control ciliary transcriptional programs during spermatogenesis.

Introduction

Cilia and flagella are evolutionarily conserved hair-like microtubule organelles that project from the surfaces of most eukaryotic cells(Derderian, Canales, & Reiter, 2023; Mitchell, 2017). They play essential roles in sensory reception, signal transduction and cell movement(Goetz & Anderson, 2010; Nachury & Mick, 2019; Singla & Reiter, 2006). The dysfunction of cilia in humans leads to an emerging class of genetic diseases called ciliopathies classified into first-order and second-order ciliopathies(Ishikawa & Marshall, 2011; Lovera & Lüders, 2021; Reiter & Leroux, 2017). First-order ciliopathies are caused by aberrations in ciliary proteins, and second-order ciliopathies occur due to defects in non- ciliary proteins, such as transcription factors, that are required for cilia formation and function(Reiter & Leroux, 2017).

Cilia are complex organelles and are dynamically regulated during development and the cell cycle(Eggenschwiler & Anderson, 2007; Kasahara & Inagaki, 2021; Santos & Reiter, 2008). The transcriptional regulation of ciliary genes must be precisely coordinated during ciliogenesis(Choksi, Lauter, Swoboda, & Roy, 2014; Collins, Ventrella, & Mitchell, 2021; Lewis & Stracker, 2021; Thomas et al., 2010). In metazoans, several transcription factors, including FOXJ1 and RFXs, have been shown to be involved in directing the expression of ciliary genes(Stubbs, Oishi, Izpisua Belmonte, & Kintner, 2008; Swoboda, Adler, & Thomas, 2000; Yu, Ng, Habacher, & Roy, 2008). These key ciliary transcription factors show dynamic expression patterns during development and differentiation(Stubbs et al., 2008; Swoboda et al., 2000; Yu et al., 2008). However, how these key transcriptional programs are spatially and temporally controlled by cell type-specific transcription factors and signaling pathways in order to activate specific target ciliary genes and generate diverse cilia is largely unknown.

Despite transcriptional regulation of ciliary genes is required for ciliogenesis in both unicellular and multicellular organisms, the underlying mechanism mediating ciliary gene expression may be fundamentally different, as the two major transcription factors, FOXJ1 and RFXs, are absent from many unicellular organisms(Chu, Baillie, & Chen, 2010; Li et al., 2018; Piasecki, Burghoorn, & Swoboda, 2010; Vij et al., 2012). Cilia and ciliary genes have been highly conserved throughout evolution, suggesting that the regulation of ciliary genes could be programmed by yet undiscovered transcriptional mechanisms, which possibly coevolved with multicellularity. Recently, an ancient transcription factor, XAP5, is found to regulate ciliary gene expression in a unicellular organism, Chlamydomonas reinhardtii(Li et al., 2018). XAP5 is highly conserved among different species and is widely expressed across tissues(Martin-Tryon & Harmer, 2008; Mazzarella, Pengue, Yoon, Jones, & Schlessinger, 1997). In vertebrates, XAP5-like (XAP5L) with an intronless open reading frame originated in Therians via retrotransposition of the ancestral XAP5, and it is highly expressed during spermatogenesis(Sedlacek et al., 1999; A. Zhang et al., 2011). However, whether XAP5 and XAP5L play essential roles in cilia development in multicellular organisms remains unknown.

Here, we show that XAP5/XAP5L are antagonistic transcription factors required for ciliary transcriptional programs during spermatogenesis. XAP5 is widely expressed in different tissues, while XAP5L is exclusively present in the testes. To explore the roles of XAP5/XAP5L in ciliogenesis, we generated a XAP5L knockout (KO) mouse line and found that XAP5L KO male mice are infertile due to the malformation of sperm flagella. Interestingly, male mice with the loss of XAP5 protein in germ cells also exhibited infertile, due to the arrest of spermatogenesis in the meiotic prophase stage. These data suggest that XAP5/XAP5L play crucial roles in sperm flagellar assembly. Mechanistically, XAP5 positively modulates transcriptional rewiring of ciliary genes via activating the key regulators including FOXJ1 and RFXs during the early stage of spermatogenesis. Conversely, XAP5L negatively regulates the transcriptional program controlling ciliogenesis by repressing these ciliary transcription factors during the spermiogenesis stage. Thus, ciliary transcriptional programs are spatially and temporally controlled by XAP5/XAP5L antagonistic transcription factors during spermatogenesis.

Results

XAP5 and XAP5L are indispensable for male fertility

To explore the potential functions of XAP5/XAP5L during ciliogenesis in vivo, we performed Western blot analysis to investigate the spatiotemporal expression of XAP5/XAP5L in various mouse tissues. We observed that XAP5 protein was widely expressed in diverse tissues and present at all stages of postnatal testes maturation, whereas XAP5L protein was mainly present in the testes and its level increased dramatically from postnatal day 21 (P21) and continued into adulthood (Figures 1A, B). Next, we identified the specific testicular cell types expressing XAP5/XAP5L using published single-cell RNA-seq data(Jung et al., 2019). XAP5 mRNA was found predominantly expressed in spermatogonia, while XAP5L mRNA was observed in pachytene spermatocytes, and remained until elongating spermatids (Figure 1C). Consistent with the mRNA localization patterns, immunofluorescence staining showed that XAP5 protein was mainly detected in the nuclei of spermatogonia, whereas XAP5L protein was present in the nuclei of pachytene spermatocytes and spermatids within the seminiferous tubules, and neither of XAP5 or XAP5L was detected in mature sperm (Figure 1D and Figure S1).

Loss of XAP5 and XAP5L proteins results in male mice infertility.

(A) Detection of XAP5/XAP5L protein in different tissues of adult mice by Western blot analysis. (B) Expression of XAP5/XAP5L protein during postnatal testicular development. β-actin was used as control. (C) t-SNE plots displaying XAP5/XAP5L gene expression levels in individual cell types of mouse testis visualized using published single-cell data(Jung et al., 2019). Black arrow represents the developmental pseudotime corresponding to the developmental ordering of each cell progressing through spermatogenesis. Dot color intensity represents the expression level. Cells from different stages are color-coded. (D) Immunofluorescence staining of XAP5/XAP5L in adult WT testis. Representative spermatogonium (Sg), round spermatid (Rs) and elongating spermatid (Es) are indicated. Scale bars: 20 μm (white), 5 μm (red). (E) Similar development between WT and XAP5L KO mice. Scale bar: 1 cm. (F) Average litter size of pups obtained by fertility test. Each male was caged with two females for 2 months, 5 males were used for each genotype. ***P < 0.001, n.s. stands for not significant. (G) Similar appearance of WT and XAP5 cKO mice. Scale bar: 1 cm. (H) Average litter size of pups obtained by fertility test. Each male was caged with two WT females for 2 months, 4 males were used for each genotype. ***P < 0.001. Error bars depict means ± SEM. All P-values were calculated using an unpaired, two-tailed Student’s t-test.

To investigate the physiological function of the uncharacterized XAP5/XAP5L protein, we attempted to generate global KO mice using the CRISPR-Cas9 system. We successfully established a XAP5L KO mouse line (Figures S2A-C) and found that while all XAP5L KO mice survived to adulthood and appeared indistinguishable from the wildtype (WT) mice, XAP5L KO males were sterile (Figures 1E, F). Notably, generating mosaic F0 founders with XAP5 mutant allele proved challenging, and no F1 generation inherited the mutant alleles from the few F0 founders in over a year of breeding experiments, underscoring the critical roles of XAP5 in mouse development. Subsequently, we generated floxed-XAP5 mice and crossed them with Stra8-GFPCre knockin mice(Lin et al., 2017) to produce XAP5 germline-specific KO mice (XAP5 cKO) (Figures S2D-F). Similar in appearance to controls (XAP5fl/Y), all XAP5 cKO males were also found to be sterile (Figures 1G, H). Collectively, these results suggest that XAP5/XAP5L play essential roles in spermatogenesis.

XAP5 and XAP5L function at different stages of spermatogenesis

To uncover the cause of male infertility, we conducted gross and histological analyses of the testes from WT and XAP5L KO mice. There were no differences in the gross appearance of testis or in the weight of mouse body and testis between WT and XAP5L KO mice (Figure 2A). Histological analysis did not reveal any severe abnormity in the seminiferous epithelial cycle of XAP5L KO testes (Figure 2B). We then investigated the motility of sperm collected from cauda epididymis, and found that total motility of sperm was significantly reduced in XAP5L KO mice (Figure 2C and Videos S1, 2). Further examination of the morphology and structure of cauda epididymal sperm in XAP5L KO mice revealed significantly higher rates of abnormal flagella (Figures 2D-F), indicating that XAP5L is crucial for flagellar assembly during spermiogenesis.

XAP5L is required for normal sperm formation.

(A) Similar gross testes morphology, body weight and testis weight between WT and XAP5L KO mice. Scale bar: 1 mm. n = 6, n.s. stands for not significant. (B) H&E staining showing similar histological structure of the testes from WT and XAP5L KO mice. Scale bars: 100 μm. (C) Significantly reduced sperm total motility of XAP5L KO mice. n = 3, **P < 0.01. (D) H&E staining showing the morphology of sperm in WT and XAP5L KO mice. Scale bars: 10 μm. (E) Significantly increased abnormal sperm in XAP5L KO mice. n = 6, ***P < 0.001. (F) Electron microscopic analysis displaying ultrastructural defects in XAP5L KO sperm flagella, including cytoplasmic vacuoles enveloping mitochondrial sheath (as shown by red arrowheads), a complete lack of mitochondrial sheath (as shown by red dotted circle), two or more cross-sections of the same sperm flagellum enclosed within one cell membrane (as shown by black arrowheads), partially formed outer dense fibers, and the axonemal microtubules (as indicated by asterisks). Scale bars: 2 μm. Error bars depict means ± SEM. All P-values were calculated using an unpaired, two-tailed Student’s t-test.

We also investigated the spermatogenic defects in the seminiferous tubules and epididymis of XAP5 cKO male mice. Notably, the testes of XAP5 cKO males were significantly smaller compared to WT controls at P16 and older (Figure 3A). Many seminiferous tubules in P16 and older XAP5 cKO males were vacuolated due to the absence of spermatocytes and spermatids, as indicated by the lack of XY body formation and the absence of PNA signal, a marker of spermatids and sperm, in XAP5 cKO germ cells (Figures 3B-D). As a consequence, no sperm were observed in the adult cauda epididymides of XAP5 cKO mice (Figure 3E). Taken together, these data indicate that germ cell development in XAP5 cKO males was arrested in meiotic prophase, and XAP5 is indispensable for meiotic progression during spermatogenesis.

Ablation of XAP5 in germ cells results in arrested spermatogenesis and absence of sperm.

(A) Testes sizes, body weight and testis weight from WT and XAP5 cKO mice at various ages were evaluated. Similar body weight was observed between WT and XAP5 cKO mice. No significant difference in size and weight was detected between WT and XAP5 cKO testes until P16. Scale bars: 1 mm. n = 5, ***P < 0.001, n.s. stands for not significant. (B) Histology of testes from WT and XAP5 cKO mice at various ages were evaluated by H&E staining. No pachytene spermatocytes were found, and many vacuolated tubules were observed in XAP5 cKO testis since P16. Black arrowheads indicate pachytene spermatocytes; black arrows indicate round spermatids. Vacuolated seminiferous tubules are indicated by asterisks. Scale bars: 20 μm (black), 2.5 μm (red). (C) Chromosome spreads staining showing pachytene spermatocytes from P16 WT male mouse. Yellow dotted circle indicates the XY body. Scale bars: 10 μm. (D) Immunofluorescence staining of PNA in adult testes indicating the absence of spermatids in XAP5 cKO mice. Scale bars: 20 μm. (E) H&E staining of adult epididymis displaying absence of sperm in XAP5 cKO mice. Scale bars: 50 μm. Error bars depict means ± SEM. All P-values were calculated using an unpaired, two-tailed Student’s t-test.

XAP5 and XAP5L function as antagonistic regulators of ciliary transcriptional regulatory networks via regulating transcription factors

To gain insight into the regulatory mechanisms of XAP5L during sperm flagellar assembly, we employed RNA-seq analysis to examine mature sperm collected from the cauda epididymis of WT and XAP5L KO adult mice. Differential expression analysis identified 2093 upregulated and 267 downregulated genes in XAP5L KO sperm (Figure S3A). Gene ontology analysis revealed that the upregulated genes were associated with cilia assembly or function (Figure 4A and Figures S3B, C). When crossed with the previously compiled ciliogenesis-related gene list(Nemajerova et al., 2016), the upregulated genes were observed encompassing diverse structural and functional components of the ciliogenic program (Table S1 and Figure S3D). Notably, several key transcription factor (TF) regulators of ciliogenesis were among the upregulated genes, including FOXJ1 and RFX families (Figures 4B, C and Figure S3E). Moreover, several key regulators of spermatogenesis, particularly spermiogenesis, were identified, including RFX2(Kistler et al., 2015; Wu et al., 2016), SOX families(Schartl et al., 2018; D. Zhang et al., 2018), TAF7L(Zhou et al., 2013) and TBPL1(Martianov et al., 2001) (Figure 4B and Figure S3E). Many core ciliary genes were also upregulated, such as CFAP206(Beckers et al., 2020) and IFT81(Qu et al., 2020) which are critical for male fertility (Figure S3E), consistent with the tight coregulation between spermiogenesis and cilia-related genes during spermatogenesis. These findings suggest that the genes identified in our study provide an excellent resource for candidates with novel ciliary or spermatogenesis-related functions, and the newly identified testes-specific protein TULP2(Oyama et al., 2022; Zheng et al., 2021) was selected for validation. We observed that TULP2 KO male mice were sterile due to malformation of sperm flagella (Figure S4). Overall, our results clearly indicate that XAP5L acts as a central transcriptional repressor of ciliogenesis during mouse spermatogenesis.

Aberrant ciliogenic and spermatogenic gene expression in XAP5L KO and XAP5 cKO male germ cells.

(A) Gene ontology analysis showing top ten biological process terms significantly enriched among upregulated genes in XAP5L KO sperm. (B) Real-time PCR validation of genes involved in ciliogenesis and spermatogenesis. n = 3, **P < 0.01, ***P < 0.001, n.s. stands for not significant. (C) Detection of FOXJ1 and RFX2 in mature sperm of WT and XAP5L KO mice by Western blot analysis. GAPDH was used as a control. (D) Gene ontology analysis showing top ten biological process terms significantly enriched among downregulated genes in P16 XAP5 cKO testes. (E) Real-time PCR validation of genes involved in ciliogenesis and spermatogenesis. n = 3, *P < 0.05, **P < 0.01, ***P < 0.001, n.s. stands for not significant. (F) Detection of FOXJ1 and RFX2 in P16 testes of WT and XAP5 cKO mice by Western blot analysis. GAPDH was used as a control. (G) A model on the involvement of XAP5 and XAP5L in the regulation of ciliogenesis during spermatogenesis. Error bars depict means ± SEM. All P-values were calculated using an unpaired, two-tailed Student’s t-test.

Furthermore, we performed RNA-seq analysis using P16 testes samples from WT and XAP5 cKO mice to explore the underlying mechanism of XAP5-mediated germ cell loss. Differential expression analysis identified 554 upregulated and 1587 downregulated genes in XAP5 cKO sperm (Figure S5A). We observed that the downregulated genes in XAP5 cKO males were involved in cilium assembly or functions (Figure 4D and Figures S5B-D). Strikingly, many of these downregulated genes were upregulated in XAP5L KO sperm, including the key ciliogenesis regulators FOXJ1 and RFX families (Figures 4E, F, Figure S5E and Table S1). In addition, meiosis initiation marker STRA8(Anderson et al., 2008; Ferder et al., 2019), meiosis-specific genes SPO11(Romanienko & Camerini-Otero, 2000) and DMC1(Bishop, Park, Xu, & Kleckner, 1992), key transcription regulators of spermiogenesis RFX2(Kistler et al., 2015; Wu et al., 2016), SOX30(D. Zhang et al., 2018) and CREM(Blendy, Kaestner, Weinbauer, Nieschlag, & Schutz, 1996; Nantel et al., 1996) were also downregulated in XAP5 cKO mice (Figures 4E and S5F), consistent with the aforementioned result that loss of XAP5 induced spermatogenesis arrest in meiotic stage. Considering the evolutionary conservation of XAP5/XAP5L protein across different species and the expression patterns of XAP5, XAP5L, FOXJ1 and RFX factors during spermatogenesis (Figures 1C and S6), these results suggest that XAP5 and XAP5L have antagonistic effects, and they may function upstream of the transcription factor families, including FOXJ1 and RFX factors, to coordinate the ciliogenesis during spermatogenesis (Figure 4G).

Discussion

Cilia and flagella are ancient organelles present in the last eukaryotic common ancestor (LECA)(Mitchell, 2017). Cilia assembly and maintenance are under strict transcriptional regulation in both unicellular and multicellular organisms(Choksi et al., 2014; Collins et al., 2021; Lewis & Stracker, 2021; Thomas et al., 2010). Despite cilia and flagella having an ancient origin, the evolutionary history of ciliary gene regulation has remained an unsolved problem. The main reason is that the master ciliary transcription factors found in multicellular organisms, including FOXJ1 and RFXs, are absent from unicellular organisms(Choksi et al., 2014; Collins et al., 2021; Lewis & Stracker, 2021; Thomas et al., 2010). Previously, we showed that an ancient transcription factor, XAP5, was required for flagella assembly in the unicellular green algae Chlamydomonas(Li et al., 2018). XAP5 proteins are evolutionarily conserved across diverse organisms(Li et al., 2018; Martin- Tryon & Harmer, 2008), which offers the possibility to investigate the conservation of the role of XAP5 in multicellular organisms. In the present study, we report that XAP5 positively regulates transcriptional network of ciliary genes by activating the key regulators including FOXJ1 and RFXs during spermatogenesis.

Cilia are dynamic organelles, and the expression profile of ciliary genes is dynamic during the cell cycle and under certain physiological conditions(Eggenschwiler & Anderson, 2007; Kasahara & Inagaki, 2021; Santos & Reiter, 2008). Cells have developed regulatory mechanisms to generate functional cilia in a temporal and spatial manner(Choksi et al., 2014; Collins et al., 2021; Lewis & Stracker, 2021; Thomas et al., 2010). However, how ciliary transcription factors determine temporal and spatial ciliary transcriptional programs is poorly understood. Sperm flagellar assembly is likewise tightly regulated during a highly complex temporal process named spermatogenesis(Holstein, Schulze, & Davidoff, 2003; Thomas et al., 2010). Several key transcriptional regulators of spermatogenesis, such as FOXJ1 and RFX2, have been identified(Chen, Knowles, Hebert, & Hackett, 1998; Kistler et al., 2015; Wu et al., 2016). The flagellar formation of FOXJ1-null sperm is severely impaired, and the FOXJ1 target genes are required for sperm motility(Beckers et al., 2020; Chen et al., 1998; Weidemann et al., 2016). RFX2 KO mice are completely sterile and RFX2 regulates the expression of ciliary genes during spermiogenesis(Kistler et al., 2015; Wu et al., 2016). Ciliary transcription factors (FOXJ1 and RFX2) are dynamically regulated during spermatogenesis(Jung et al., 2019). Intriguingly, our results show that XAP5 and XAP5L are two conserved pairs of antagonistic transcription factors that regulate these key transcriptional regulators required for spermatogenesis.

By using Western blot analysis, we found that XAP5 was ubiquitously expressed in all tissues examined, whereas XAP5L expression was restricted to the testes (Fig. 1a).

Interestingly, previous research found that XAP5L was widely expressed in normal tissue via using TCGA expression data(Thompson et al., 2021). Moreover, loss of XAP5/XAP5L perturbs transcriptional programs in cancer cells. At present, as shown by the recent data, we cannot rule out the possibility that XAP5L is dynamically expressed in other tissues during development. Although XAP5 and XAP5L are highly expressed in testes and critical for spermatogenesis, they are not present in mature sperm (Extended Data Fig. 1). Thus, XAP5/XAP5L are not needed for sperm maintenance. Given the dynamic expression of XAP5/XAP5L during spermatogenesis, it will be interesting to identify the key signaling factors that regulate the timely and spatial expression of XAP5 and XAP5L.

A role for XAP5 in human brain development is suggested by the association of X-linked intellectual disability (XLID) with rare XAP5 missense variants(Lee et al., 2020). However, how defects in XAP5 lead to XLID syndrome is unknown. Although there is as yet no evidence that XAP5 regulates ciliogenesis in any system besides Chlamydomonas and mice, the function in cilia could help clarify the unexplained role of XAP5 mutations in causing XLID. The possibility that XAP5 could be a ciliary transcription factor in humans would add XLID to the growing list of the second-order ciliopathies.

Materials and Methods

Mice

C57BL/6J strains (from Vital River Laboratories, Beijing, China) and Stra8-GFPCre knockin mice (from Cyagen Biosciences) were used. Floxed-XAP5 mice, XAP5L and TULP2 knockout mice were generated on the C57BL/6J background using the CRISPR/Cas9 system. The sgRNA sequences used to target XAP5, XAP5L and TULP2 were as follows: XAP5L, sgRNA1-GGC TAC CAG AAA CAG GGA CT, sgRNA2-GGG AGT AAG GTC CCC AAA CT; XAP5, sgRNA3-CTA CAG GGC ACT TAT TAA TA, sgRNA4-ATA GTA ATT CCC CCG TGC TT; TULP2, sgRNA5-TGA CTA ATT AGG CCC GAG AG, sgRNA6-GGT TCT TAG AGA GTC AAC GT. Experimental protocols were approved by the ethics committees of Jianghan University. Mice were maintained in a pathogen-free environment with a room temperature of 23 ± 2 °C under a humidity level of 30-70%. The light was maintained on a 12-hour day/night cycle.

Genome PCR

Mouse genomic DNA was isolated from the tail tip using One Step Mouse Genotyping Kit (Vazyme, PD101), and subjected to PCR with 2 × Taq Plus Master Mix (Vazyme, P212) following the manufacturer’s instructions. The PCR products were separated by 2% agarose gel electrophoresis. The primers used are listed in Table S2.

RNA isolation and real-time PCR

Total RNA was isolated using the TRIzol reagent (Invitrogen, 15596018) and then converted to cDNA with HiScript II 1st Strand cDNA Synthesis Kit (Vazyme, R212) following the manufacturer’s instructions. Real-time PCR was conducted using ChamQ SYBR qPCR Master Mix (Vazyme, Q311) on an ABI StepOnePlus real-time PCR system (Applied Biosystems). Expression values were normalized using the ΔΔCt method with Gapdh as the internal control. The primer sequences are indicated in Table S2.

Western blot analysis

Cell lysates from mouse tissues were prepared in RIPA buffer (Beyotime, P0013B) containing Protease Inhibitor Cocktail (Roche, 11873580001) and 1mM PMSF. Subcellular lysates of testicular nuclear and cytoplasmic fractions were prepared using Nuclear and Cytoplasmic Protein Extraction Kit (Beyotime, P0027) following the manufacturer’s instructions. The protein lysates were subjected to standard immunoblotting analysis. Antibodies used were as follows: anti-β-actin (ABclonal, AC026), anti-XAP5 (Prospertech, HU-412003), anti-XAP5 (Sigma, HPA003585), anti- XAP5L (Prospertech, Hu-412001), anti-LMNB (Proteintech, 66095-1-Ig), anti-GAPDH (ABclonal, AC002), anti-FOXJ1 (D360355, Sangon Biotech), anti-RFX2 (K110716P, Solarbio) and anti-TULP2 (Prospertech, Hu-412002).

Immunofluorescence

Mouse testis were fixed in 4% paraformaldehyde in PBS at 4 °C overnight, dehydrated in gradient ethanol, embedded in paraffin, and sectioned into 5 µm slices. After deparaffinization and rehydration, testis sections were boiled in 1 mM EDTA, pH 8.0 for 15 min using a microwave oven for antigen retrieval, followed by cooling to room temperature. The primary antibodies, including anti-XAP5 (Prospertech, HU-412003, 1:250 dilution), anti-XAP5 (Sigma, HPA003585, 1:250 dilution), anti-XAP5L (Prospertech, Hu-412001, 1:250 dilution), anti-lectin PNA, Alexa Fluor 488 Conjugate (ThermoFisher, L21409, 1:250 dilution), and anti-TULP2 (Prospertech, Hu-412002, 1:250 dilution), were then incubated at 4 °C overnight. Subsequently, the sections were incubated with the secondary antibody donkey anti-rabbit Alexa Fluor Plus 488 (Invitrogen, A32790, 1:500 dilution) for 1 hour and DAPI for 5 minutes at room temperature. Images were collected using a Zeiss Axio Vert A1 microscope.

Histological analysis

Testes and epididymides were fixed in Bouin’s solution (sigma, HT10132) overnight at 4 °C, embedded in paraffin and sectioned into 5 µm slices. The sections were then subjected to H&E staining using standard procedures and observed under a Zeiss Axio Vert A1 microscope.

Assessment of sperm motility and morphology

Cauda epididymides were collected from male mice and dissected in 37°C pre-warmed human tubal fluid (HTF) culture medium (EasyCheck, M1130), followed by incubation at 37°C for 15 min to release the sperm. To assess sperm motility, the sperm suspension was observed and recorded under a Leica total internal reflection fluorescence microscope. Motility was defined as any movement of the sperm flagellum during a 10-second observation cycle. After immobilization with 1% paraformaldehyde, sperm morphology was observed by H&E staining using standard procedures. Images were captured using a Zeiss Axio Vert A1 microscope equipped with a ×100 oil immersion objective. Sperm without staining were also used to assess sperm morphology. For ultrastructural observation, sperm samples were fixed in 3% glutaraldehyde in 0.1M sodium phosphate buffer (pH7.4), and post-fixed with 1% (wt/vol) OsO4. After dehydration, the samples were placed in propylene oxide and embedded in a mixture of Epon 812 and Araldite. Ultrathin sections obtained by a Leica EM UC7 ultramicrotome were stained with uranyl acetate and lead citrate, then analyzed using a HT7700 TEM (Hitachi).

Fertility test

Each of the adult male mice was mated with two females for 2 months. All the mice used were aged 6-8 weeks. Vaginal plugs were checked every morning, and the number of newborn pups was counted.

Chromosome spreads and immunofluorescence

Testicular chromosome spreads were prepared and immunolabeled as described(Reinholdt, Ashley, Schimenti, & Shima, 2004). The primary antibodies used were as follows: anti-γ H2AX (Millipore, 05-636, 1:300 dilution), anti-SYCP3 (Proteintech, 23024-1-AP, 1:300 dilution). Secondary antibodies used were Alexa Dye (AlexaFluor Plus 488/594) conjugates (ThermoFisher) at 1:500 dilutions. Images were collected with a Zeiss Axio Vert A1 microscope.

RNA-seq analysis

The mature sperm samples from P56 mice and testis samples from P16 mice were washed three time in 1× PBS after harvest. Total RNA was extracted from the samples using the TRIzol reagent (Invitrogen) and purified with the NEBNext Poly(A) mRNA Magnetic Isolation Module (NEB, E7490). The quality of the RNA samples was examined using the Agilent 2100 Bioanalyzer system (Agilent Technologies). Qualified RNA samples were subjected to sequencing library construction using the NEBNext Ultra II RNA Library Prep Kit for Illumina (NEB, E7775). Paired-end (2 × 150 bp) sequencing was carried out on the Illumina NovaSeq6000 platform. The sequencing reads were mapped to the Mus musculus reference genome (GRCm38/mm10) using HISAT2 v2.1.0(Kim, Langmead, & Salzberg, 2015), and StringTie v1.3.3b(Pertea et al., 2015) was applied to assemble the mapped reads. Fragments Per Kilobase of transcript per Million fragments mapped (FPKM) was used to measure the expression level of a gene. Differential expression analysis was processed by DESeq2 v1.12.4(Subramanian et al., 2005), and the significant differentially expressed genes were identified with a fold change >2. Gene Ontology analysis was performed using the DAVID database.

Statistical analysis

Statistical analyses were conducted using GraphPad Prism 8.0.2 software. All data were presented as mean ± SEM. Values of P < 0.05 were considered statistically significant.

Statistical significance between two groups was calculated using an unpaired, parametric, two-sided student’s t test.

Supporting information

eLife Assessment

This study reports useful data suggesting the critical roles of two ancient proteins, XAP5 and XAP5L, in controlling the transcriptional program of ciliogenesis during mouse spermatogenesis. However, this study is considered incomplete because the data only partially support the conclusion. This work will be of interest to biomedical researchers who work on ciliogenesis and reproduction.

Reviewer #1 (Public Review):

Summary:

Wang et al. generate XAP5 and XAP5L knockout mice and find that they are male infertile due to meiotic arrest and reduced sperm motility, respectively. RNA-Seq was subsequently performed and the authors concluded that XAP5 and XAP5L are antagonistic transcription factors of cilliogenesis (in XAP5-KO P16 testis: 554 genes were unregulated and 1587 genes were downregulated; in XAP5L-KO sperm: 2093 genes were unregulated and 267 genes were downregulated).

Strengths:

Knockout mouse models provided strong evidence to indicate that XAP5 and XAP5L are critical for spermatogenesis and male fertility.

Weaknesses:

The key conclusions are not supported by evidence. First, the authors claim that XAP5 and XAP5L transcriptionally regulate sperm flagella development; however, detailed molecular experiments related to transcription regulation are lacking. How do XAP5 and XAP5L regulate their targets? Only RNA-Seq is not enough. Second, the authors declare that XAP5 and XAP5L are antagonistic transcription factors; however, how do XAP5 and XAP5L regulate sperm flagella development antagonistically? Only RNA-Seq is not enough. Third, I am concerned about whether XAP5 really regulates sperm flagella development. XAP5 is specifically expressed in spermatogonia and XAP5-cKO mice are in meiotic arrest, indicating that XAP5 regulates meiosis rather than sperm flagella development.

Reviewer #2 (Public Review):

In this study, Wang et al., report the significance of XAP5L and XAP5 in spermatogenesis, involved in transcriptional regulation of the ciliary gene in testes. In previous studies, the authors demonstrate that XAP5 is a transcription factor required for flagellar assembly in Chlamydomonas. Continuing from their previous study, the authors examine the conserved role of the XAP5 and XAP5L, which are the orthologue pair in mammals.

XAP5 and XAP5L express ubiquitously and testis specifically, respectively, and their absence in the testes causes male infertility with defective spermatogenesis. Interestingly, XAP5 deficiency arrests germ cell development at the pachytene stage, whereas XAP5L absence causes impaired flagellar formation. RNA-seq analyses demonstrated that XAP5 deficiency suppresses ciliary gene expression including Foxj1 and Rfx family genes in early testis. By contrast, XAP5L deficiency abnormally remains Foxj1 and Rfx genes in mature sperm. From the results, the authors conclude that XAP5 and XAP5L are the antagonistic transcription factors that function upstream of Foxj1 and Rfx family genes.

This reviewer thinks the overall experiments are performed well and that the manuscript is clear. However, the current results do not directly support the authors' conclusion. For example, the transcriptional function of XAP5 and XAP5L requires more evidence. In addition, this reviewer wonders about the conserved XAP5 function of ciliary/flagellar gene transcription in mammals - the gene is ubiquitously expressed despite its functional importance in flagellar assembly in Chlamydomonas. Thus, this reviewer thinks authors are required to show more direct evidence to clearly support their conclusion with more descriptions of its role in ciliary/flagellar assembly.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

Wang et al. generate XAP5 and XAP5L knockout mice and find that they are male infertile due to meiotic arrest and reduced sperm motility, respectively. RNA-Seq was subsequently performed and the authors concluded that XAP5 and XAP5L are antagonistic transcription factors of cilliogenesis (in XAP5-KO P16 testis: 554 genes were unregulated and 1587 genes were downregulated; in XAP5L-KO sperm: 2093 genes were unregulated and 267 genes were downregulated).

We are grateful for the comprehensive summary.

Strengths:

Knockout mouse models provided strong evidence to indicate that XAP5 and XAP5L are critical for spermatogenesis and male fertility.

Thank you for your positive comment.

Weaknesses:

The key conclusions are not supported by evidence. First, the authors claim that XAP5 and XAP5L transcriptionally regulate sperm flagella development; however, detailed molecular experiments related to transcription regulation are lacking. How do XAP5 and XAP5L regulate their targets? Only RNA-Seq is not enough. Second, the authors declare that XAP5 and XAP5L are antagonistic transcription factors; however, how do XAP5 and XAP5L regulate sperm flagella development antagonistically? Only RNA-Seq is not enough. Third, I am concerned about whether XAP5 really regulates sperm flagella development. XAP5 is specifically expressed in spermatogonia and XAP5-cKO mice are in meiotic arrest, indicating that XAP5 regulates meiosis rather than sperm flagella development.

Thank you for the critical comments. To strengthen our conclusions, we have included XAP5/XAP5L CUT&Tag data in our revised manuscript. This highly sensitive method has allowed us to identify direct target genes of XAP5 and XAP5L (Table S1, Figure S6). Notably, our results demonstrate that both FOXJ1 and RFX2 are occupied by XAP5 (Figure 4G). Additionally, real-time PCR validation confirmed that RFX2 is also associated with XAP5L, even though enriched peaks for the RFX2 gene were not detected in the initial CUT&Tag data (Figure 4G). These findings indicate that XAP5 and XAP5L regulate the expression of FOXJ1 and RFX2 by directly binding to these genes. De novo motif analyses revealed that XAP5 and XAP5L shared a conserved binding sequence (CCCCGCCC/GGGCGGGG) (Figure S6C), and the bound regions of FOXJ1 and RFX2 contain this sequence. Further analysis shows that many XAP5L target genes are also targets of XAP5 (Figure S6G), despite the limited number of identified XAP5L target genes. This differential binding and regulation of shared target genes underscore the antagonistic relationship between XAP5 and XAP5L. Collectively, these findings provide additional support for the idea that XAP5 and XAP5L function as antagonistic transcription factors, acting upstream of transcription factor families, including FOXJ1 and RFX factors, to coordinate ciliogenesis during spermatogenesis.

While we agree that XAP5 primarily regulates meiosis during spermatogenesis, our data also indicate that many cilia-related genes, including key transcription regulators of spermiogenesis such as RFX2 and SOX30, are downregulated in XAP5-cKO mice and are bound by XAP5 (Figure 4, Figures S4 and S6). It is important to note that genes coding for flagella components are expressed sequentially and in a germ cell-specific manner during development. When we refer to "regulating sperm flagella development", we mean the spatiotemporal regulation. We have revised the manuscript to clarify this point.

Reviewer #2 (Public Review):

In this study, Wang et al., report the significance of XAP5L and XAP5 in spermatogenesis, involved in transcriptional regulation of the ciliary gene in testes. In previous studies, the authors demonstrate that XAP5 is a transcription factor required for flagellar assembly in Chlamydomonas. Continuing from their previous study, the authors examine the conserved role of the XAP5 and XAP5L, which are the orthologue pair in mammals.

XAP5 and XAP5L express ubiquitously and testis specifically, respectively, and their absence in the testes causes male infertility with defective spermatogenesis. Interestingly, XAP5 deficiency arrests germ cell development at the pachytene stage, whereas XAP5L absence causes impaired flagellar formation. RNA-seq analyses demonstrated that XAP5 deficiency suppresses ciliary gene expression including Foxj1 and Rfx family genes in early testis. By contrast, XAP5L deficiency abnormally remains Foxj1 and Rfx genes in mature sperm. From the results, the authors conclude that XAP5 and XAP5L are the antagonistic transcription factors that function upstream of Foxj1 and Rfx family genes.

This reviewer thinks the overall experiments are performed well and that the manuscript is clear. However, the current results do not directly support the authors' conclusion. For example, the transcriptional function of XAP5 and XAP5L requires more evidence. In addition, this reviewer wonders about the conserved XAP5 function of ciliary/flagellar gene transcription in mammals - the gene is ubiquitously expressed despite its functional importance in flagellar assembly in Chlamydomonas. Thus, this reviewer thinks authors are required to show more direct evidence to clearly support their conclusion with more descriptions of its role in ciliary/flagellar assembly.

Thank you for your thoughtful review of our work. We appreciate your positive feedback on the overall quality of the experiments and the clarity of the manuscript. In response to your concerns, we have included new experimental data and made revisions to the manuscript (lines 193-217) to better support our conclusions, particularly regarding the transcriptional function of XAP5 and XAP5L. Additionally, we have expanded on the role of XAP5 in ciliary and flagellar assembly to provide more direct evidence for its functional importance. Thank you for your insights.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

The title (Control of ciliary transcriptional programs during spermatogenesis by antagonistic transcription factors) is not specific and does tend to exaggerate.

Thank you for the comment, and we appreciate the opportunity to clarify the appropriateness of the title. Our paper extensively investigates the transcriptional regulation of ciliary genes during spermatogenesis. It demonstrates that XAP5/XAP5L are key transcription factors involved in this process. The title reflects our primary focus on the transcriptional programs that govern ciliary gene expression. Moreover, our paper shows that XAP5 positively regulates the expression of ciliary genes, particularly during the early stages of spermatogenesis, while XAP5L negatively regulates these genes. This antagonistic relationship is a crucial aspect of the study and is effectively conveyed in the title. In addition, our revised paper provides detailed insights into how XAP5/XAP5L control ciliary gene expression during spermatogenesis.

Figure 4C: FOXJ1 and RFX2 are absent in sperm from WT mice. Are you sure? They are highly expressed in WT testes.

Thank you for your careful review. While FOXJ1 and RFX2 are indeed highly expressed in the testes of wild-type (WT) mice, our data show that they are not detectable in mature sperm. This observation is consistent with published single-cell RNA-seq data(Jung et al., 2019), which indicate that FOXJ1 and RFX2 are primarily expressed in spermatocytes but not in spermatids (Figure S7). This expression pattern aligns with that that of IFT-particle proteins, which are essential for the formation but not the maintenance of mammalian sperm flagella(San Agustin, Pazour, & Witman, 2015).

XAP5 is specifically expressed in spermatogonia and XAP5-cKO mice are in meiotic arrest, indicating that XAP5 regulates meiosis rather than sperm flagella development.

We appreciate your insightful comments. As mentioned above, we agree that XAP5 primarily regulates meiosis during spermatogenesis. When we mentioned "regulating sperm flagella development," we were referring to the spatiotemporal regulation of these processes. We have revised the manuscript to clarify this distinction. Thank you for your understanding.

The title of Figure 2 (XAP5L is required for normal sperm formation) is not accurate because the progress of spermatogenesis and sperm count is normal in XAP5L-KO mice (only sperm motility is reduced).

We apologize for any confusion caused by the previous figure. It did not accurately convey the changes in sperm count. In the revised Figure 2B, we clearly demonstrate that the sperm count in XAP5L-KO mice is indeed lower than that in WT mice. This revision aims to provide a more accurate representation of the effects of XAP5L deficiency on spermatogenesis. Thank you for bringing this to our attention.

Reviewer #2 (Recommendations For The Authors):

(1) Although XAP5 and XAP5L deficiency alters the transcription of Foxj1 and Rfx family genes, which are the essential transcription factors for the ciliogenesis, current data do not directly support that XAP5 and XAP5L are the upstream transcription factors. The authors need to show more direct evidence such as CHIP-Seq data.

Thank you for your valuable feedback! In this revised manuscript, we have included data identifying candidate direct targets of XAP5 and XAP5L using the highly sensitive CUT&Tag method (Kaya-Okur et al., 2019). Our results show that XAP5 occupies both FOXJ1 and RFX2 (Figure 4G). Furthermore, real-time PCR validation of the CUT&Tag experiments confirmed that RFX2 is also occupied by XAP5L (Figure 4G), despite the initial CUT&Tag data not revealing enriched peaks for the RFX2 gene (Table S1). Unfortunately, the limited number of enriched peaks identified for XAP5L (Table S1) suggests that the XAP5L antibody used in the CUT&Tag experiment might have suboptimal performance, which prevented us from detecting occupancy on the FOXJ1 promoter. Nevertheless, these additional data provide strong evidence that XAP5 and XAP5L function as upstream transcription factors for FOXJ1 and RFX family genes, supporting their essential roles in ciliogenesis.

(2) Shared transcripts that are altered by the absence of either XAP5 or XAP5L do not clearly support they are antagonistic transcription factors.

Thank you for your insightful comment. In our revised manuscript, we performed CUT&Tag analysis to identify target genes of XAP5 and XAP5L. Motif enrichment analysis revealed conserved binding sequences for both factors (Figures S6C), indicating a subset of shared downstream genes between XAP5 and XAP5L. Among the downregulated genes in XAP5 cKO germ cells, 891 genes were bound by XAP5 (Figure S6D). Although the number of enriched peaks identified for XAP5L was limited, 75 of the upregulated genes in XAP5L KO sperm were bound by XAP5L (Figure S6E). Importantly, of these 75 XAP5L target genes, approximately 30% (22 genes) were also identified as targets of XAP5 (Figure S6G), further support the idea that XAP5 and XAP5L function as antagonistic transcription factors.

(3) XAP5 seems to be an ancient transcription factor for cilia and flagellar assembly. However, XAP5 expresses ubiquitously in mice. How can this discrepancy be explained? Is it also required for primary cilia assembly? Are their expression also directly linked to ciliogenesis in other types of cells?

Thank you for the thoughtful questions. The ubiquitous expression of XAP5 in mice can be understood in light of its role as an ancient transcription factor for cilia and flagellar assembly. Given that cilia are present on nearly every cell type in the mammalian body (O'Connor et al., 2013), this broad expression pattern makes sense. In fact, XAP5 serves not only as a master regulator of ciliogenesis but also as a critical regulator of various developmental processes (Kim et al., 2018; Lee et al., 2020; Xie et al., 2023).

Our current unpublished work demonstrates that XAP5 is essential for primary cilia assembly in different cell lines. The loss of XAP5 protein results in abnormal ciliogenesis, further supporting its vital role in ciliary formation across different cell types.

We believe that the widespread expression of XAP5 reflects its fundamental importance in multiple cellular processes, including ciliogenesis, development, and potentially other cellular functions yet to be discovered.

(4) XAP5L causes impairs flagellar assembly. Have the authors observed any other physiological defects in the absence of XAP5L in mouse models? Such as hydrocephalus and/or tracheal defects?

Thank you for the questions. We have carefully examined XAP5L KO mice for other physiological defects. To date, we have not observed any additional physiological abnormalities. Specifically, we assessed the condition of tracheal cilia in XAP5L KO mice and found no significant differences compared to wild-type (WT) mice, as illustrated in Author response image 1 below.

References

Jung, M., Wells, D., Rusch, J., Ahmad, S., Marchini, J., Myers, S. R., & Conrad, D. F. (2019). Unified single-cell analysis of testis gene regulation and pathology in five mouse strains. Elife, 8. doi:10.7554/eLife.43966

Kaya-Okur, H. S., Wu, S. J., Codomo, C. A., Pledger, E. S., Bryson, T. D., Henikoff, J. G., . . . Henikoff, S. (2019). CUT&Tag for efficient epigenomic profiling of small samples and single cells. Nat Commun, 10(1), 1930. doi:10.1038/s41467-019-09982-5

Kim, Y., Hur, S. W., Jeong, B. C., Oh, S. H., Hwang, Y. C., Kim, S. H., & Koh, J. T. (2018). The Fam50a positively regulates ameloblast differentiation via interacting with Runx2. J Cell Physiol, 233(2), 1512-1522. doi:10.1002/jcp.26038

Lee, Y.-R., Khan, K., Armfield-Uhas, K., Srikanth, S., Thompson, N. A., Pardo, M., . . . Schwartz, C. E. (2020). Mutations in FAM50A suggest that Armfield XLID syndrome is a spliceosomopathy. Nature Communications, 11(1). doi:10.1038/s41467-020-17452-6

O'Connor, A. K., Malarkey, E. B., Berbari, N. F., Croyle, M. J., Haycraft, C. J., Bell, P. D., . . . Yoder, B. K. (2013). An inducible CiliaGFP mouse model for in vivo visualization and analysis of cilia in live tissue. Cilia, 2(1), 8. doi:10.1186/2046-2530-2-8

San Agustin, J. T., Pazour, G. J., & Witman, G. B. (2015). Intraflagellar transport is essential for mammalian spermiogenesis but is absent in mature sperm. Mol Biol Cell, 26(24), 4358-4372. doi:10.1091/mbc.E15-08-0578

Xie, X., Li, L., Tao, S., Chen, M., Fei, L., Yang, Q., . . . Chen, L. (2023). Proto-Oncogene FAM50A Can Regulate the Immune Microenvironment and Development of Hepatocellular Carcinoma In Vitro and In Vivo. Int J Mol Sci, 24(4). doi:10.3390/ijms24043217

Version 2

Abstract

Abstract

Existence of cilia in the last eukaryotic common ancestor (LECA) raises a fundamental question in biology: how the transcriptional regulation of ciliogenesis has evolved? One conceptual answer to this question is by an ancient transcription factor regulating ciliary gene expression in both unicellular and multicellular organisms, but examples of such transcription factors in eukaryotes are lacking. Previously, we showed that an ancient transcription factor XAP5 is required for flagellar assembly in Chlamydomonas. Here, we show that XAP5 and XAP5L are two conserved pairs of antagonistic transcription regulators that control ciliary transcriptional programs during spermatogenesis. Male mice lacking either XAP5 or XAP5L display infertility, as a result of meiotic prophase arrest and sperm flagella malformation, respectively. Mechanistically, XAP5 positively regulates the ciliary gene expression by activating the key regulators including FOXJ1 and RFX families during the early stage of spermatogenesis. In contrast, XAP5L negatively regulates the expression of ciliary genes via repressing these ciliary transcription factors during the spermiogenesis stage. Our results provide new insights into the mechanisms by which temporal and spatial transcription regulators are coordinated to control ciliary transcriptional programs during spermatogenesis.

Introduction

Cilia and flagella are evolutionarily conserved hair-like microtubule organelles that project from the surfaces of most eukaryotic cells(Derderian, Canales, & Reiter, 2023; Mitchell, 2017). They play essential roles in sensory reception, signal transduction and cell movement(Goetz & Anderson, 2010; Nachury & Mick, 2019; Singla & Reiter, 2006). The dysfunction of cilia in humans leads to an emerging class of genetic diseases called ciliopathies classified into first-order and second-order ciliopathies(Ishikawa & Marshall, 2011; Lovera & Lüders, 2021; Reiter & Leroux, 2017). First-order ciliopathies are caused by aberrations in ciliary proteins, and second-order ciliopathies occur due to defects in non-ciliary proteins, such as transcription factors, that are required for cilia formation and function(Reiter & Leroux, 2017).

Cilia are complex organelles and are dynamically regulated during development and the cell cycle(Eggenschwiler & Anderson, 2007; Kasahara & Inagaki, 2021; Santos & Reiter, 2008). The transcriptional regulation of ciliary genes must be precisely coordinated during ciliogenesis(Choksi, Lauter, Swoboda, & Roy, 2014; Collins, Ventrella, & Mitchell, 2021; Lewis & Stracker, 2021; Thomas et al., 2010). In metazoans, several transcription factors, including FOXJ1 and RFXs, have been shown to be involved in directing the expression of ciliary genes(Stubbs, Oishi, Izpisua Belmonte, & Kintner, 2008; Swoboda, Adler, & Thomas, 2000; X. Yu, Ng, Habacher, & Roy, 2008). These key ciliary transcription factors show dynamic expression patterns during development and differentiation(Stubbs et al., 2008; Swoboda et al., 2000; X. Yu et al., 2008). However, how these key transcriptional programs are spatially and temporally controlled by cell type-specific transcription factors and signaling pathways in order to activate specific target ciliary genes and generate diverse cilia is largely unknown.

Despite transcriptional regulation of ciliary genes is required for ciliogenesis in both unicellular and multicellular organisms, the underlying mechanism mediating ciliary gene expression may be fundamentally different, as the two major transcription factors, FOXJ1 and RFXs, are absent from many unicellular organisms(Chu, Baillie, & Chen, 2010; Li et al., 2018; Piasecki, Burghoorn, & Swoboda, 2010; Vij et al., 2012). Cilia and ciliary genes have been highly conserved throughout evolution, suggesting that the regulation of ciliary genes could be programmed by yet undiscovered transcriptional mechanisms, which possibly coevolved with multicellularity. Recently, an ancient transcription factor, XAP5, is found to regulate ciliary gene expression in a unicellular organism, Chlamydomonas reinhardtii(Li et al., 2018). XAP5 is highly conserved among different species and is widely expressed across tissues(Martin-Tryon & Harmer, 2008; Mazzarella, Pengue, Yoon, Jones, & Schlessinger, 1997). In vertebrates, XAP5-like (XAP5L) with an intronless open reading frame originated in Therians via retrotransposition of the ancestral XAP5, and it is highly expressed during spermatogenesis(Sedlacek et al., 1999; A. Zhang et al., 2011). However, whether XAP5 and XAP5L play essential roles in cilia development in multicellular organisms remains unknown.

Here, we show that XAP5/XAP5L are antagonistic transcription factors required for ciliary transcriptional programs during spermatogenesis. XAP5 is widely expressed in different tissues, while XAP5L is exclusively present in the testes. To explore the roles of XAP5/XAP5L in ciliogenesis, we generated a XAP5L knockout (KO) mouse line and found that XAP5L KO male mice are infertile due to the malformation of sperm flagella. Interestingly, male mice with the loss of XAP5 protein in germ cells also exhibited infertile, due to the arrest of spermatogenesis in the meiotic prophase stage. These data suggest that XAP5/XAP5L play crucial roles in sperm flagellar assembly in a spatiotemporal manner. Mechanistically, XAP5 positively modulates transcriptional rewiring of ciliary genes via activating the key regulators including FOXJ1 and RFXs during the early stage of spermatogenesis. Conversely, XAP5L negatively regulates the transcriptional program controlling ciliogenesis by repressing these ciliary transcription factors during the spermiogenesis stage. Thus, ciliary transcriptional programs are spatially and temporally controlled by XAP5/XAP5L antagonistic transcription factors during spermatogenesis.

Results

XAP5 and XAP5L are indispensable for male fertility

To explore the potential functions of XAP5/XAP5L during ciliogenesis in vivo, we performed Western blot analysis to investigate the spatiotemporal expression of XAP5/XAP5L in various mouse tissues. We observed that XAP5 protein was widely expressed in diverse tissues and present at all stages of postnatal testes maturation, whereas XAP5L protein was mainly present in the testes and its level increased dramatically from postnatal day 21 (P21) and continued into adulthood (Figures 1A, B). Next, we identified the specific testicular cell types expressing XAP5/XAP5L using published single-cell RNA-seq data(Jung et al., 2019). XAP5 mRNA was found predominantly expressed in spermatogonia, while XAP5L mRNA was observed in pachytene spermatocytes, and remained until elongating spermatids (Figure 1C). Consistent with the mRNA localization patterns, immunofluorescence staining showed that XAP5 protein was mainly detected in the nuclei of spermatogonia, whereas XAP5L protein was present in the nuclei of pachytene spermatocytes and spermatids within the seminiferous tubules, and neither of XAP5 or XAP5L was detected in mature sperm (Figure 1D and Figure S1).

Loss of XAP5 and XAP5L proteins results in male mice infertility.

(A) Detection of XAP5/XAP5L protein in different tissues of adult mice by Western blot analysis. (B) Expression of XAP5/XAP5L protein during postnatal testicular development. β-actin was used as control. (C) t-SNE plots displaying XAP5/XAP5L gene expression levels in individual cell types of mouse testis visualized using published single-cell data(Jung et al., 2019). Black arrow represents the developmental pseudotime corresponding to the developmental ordering of each cell progressing through spermatogenesis. Dot color intensity represents the expression level. Cells from different stages are color-coded. (D) Immunofluorescence staining of XAP5/XAP5L in adult WT testis. Representative spermatogonium (Sg), round spermatid (Rs) and elongating spermatid (Es) are indicated. Scale bars: 20 μm (white), 5 μm (red). (E) Similar development between WT and XAP5L KO mice. Scale bar: 1 cm. (F) Average litter size of pups obtained by fertility test. Each male was caged with two females for 2 months, 5 males were used for each genotype. ***P < 0.001, n.s. stands for not significant. (G) Similar appearance of WT and XAP5 cKO mice. Scale bar: 1 cm. (H) Average litter size of pups obtained by fertility test. Each male was caged with two WT females for 2 months, 4 males were used for each genotype. ***P < 0.001. Error bars depict means ± SEM. All P-values were calculated using an unpaired, two-tailed Student’s t-test.

To investigate the physiological function of the uncharacterized XAP5/XAP5L protein, we attempted to generate global KO mice using the CRISPR-Cas9 system. We successfully established a XAP5L KO mouse line (Figures S2A-C) and found that while all XAP5L KO mice survived to adulthood and appeared indistinguishable from the wildtype (WT) mice, XAP5L KO males were sterile (Figures 1E, F). Notably, generating mosaic F0 founders with XAP5 mutant allele proved challenging, and no F1 generation inherited the mutant alleles from the few F0 founders in over a year of breeding experiments, underscoring the critical roles of XAP5 in mouse development. Subsequently, we generated floxed-XAP5 mice and crossed them with Stra8-GFPCre knockin mice(Lin et al., 2017) to produce XAP5 germline-specific KO mice (XAP5 cKO) (Figures S2D-F). Similar in appearance to controls (XAP5fl/Y), all XAP5 cKO males were also found to be sterile (Figures 1G, H). Collectively, these results suggest that XAP5/XAP5L play essential roles in spermatogenesis.

XAP5 and XAP5L function at different stages of spermatogenesis

To uncover the cause of male infertility, we conducted gross and histological analyses of the testes from WT and XAP5L KO mice. There were no differences in the gross appearance of testis or in the weight of mouse body and testis between WT and XAP5L KO mice (Figure 2A), but the number of sperm collected from cauda epididymis of XAP5L KO mice was significantly lower than that from WT mice (Figure 2B), and sperm motility was significantly reduced in KO mice (Figure 2C and Videos S1, 2). Histological analysis did not reveal any severe abnormity in the seminiferous epithelial cycle of XAP5L KO testes (Figure 2D). Further examination of the morphology and structure of cauda epididymal sperm in XAP5L KO mice revealed significantly higher rates of abnormal flagella (Figures 2E-G), indicating that XAP5L is crucial for flagellar assembly during spermiogenesis.

XAP5L is required for normal sperm formation.

(A) Similar gross testes morphology, body weight and testis weight between WT and XAP5L KO mice. Scale bar: 1 mm. n = 6, n.s. stands for not significant. (B) Significantly reduced caudal epididymal sperm counts in XAP5L KO mice compared with WT males. n = 5, **P < 0.01. (C) Significantly reduced sperm total motility of XAP5L KO mice. n =3, **P < 0.01. (D) H&E staining showing similar histological structure of the testes from WT and XAP5L KO mice. Scale bars: 100 μm. (E) H&E staining showing the morphology of sperm in WT and XAP5L KO mice. Scale bars: 10 μm. (F) Significantly increased abnormal sperm in XAP5L KO mice. n = 6, ***P < 0.001. (G) Electron microscopic analysis displaying ultrastructural defects in XAP5L KO sperm flagella, including cytoplasmic vacuoles enveloping mitochondrial sheath (as shown by red arrowheads), a complete lack of mitochondrial sheath (as shown by red dotted circle), two or more cross-sections of the same sperm flagellum enclosed within one cell membrane (as shown by black arrowheads), partially formed outer dense fibers, and the axonemal microtubules (as indicated by asterisks). Scale bars: 2 μm. Error bars depict means ± SEM. All P-values were calculated using an unpaired, two-tailed Student’s t- test.

We also investigated the spermatogenic defects in the seminiferous tubules and epididymis of XAP5 cKO male mice. Notably, the testes of XAP5 cKO males were significantly smaller compared to WT controls at P16 and older (Figure 3A). Many seminiferous tubules in P16 and older XAP5 cKO males were vacuolated due to the absence of spermatocytes and spermatids, as indicated by the lack of XY body formation and the absence of PNA signal, a marker of spermatids and sperm, in XAP5 cKO germ cells (Figures 3B-D). As a consequence, no sperm were observed in the adult cauda epididymides of XAP5 cKO mice (Figure 3E). Taken together, these data indicate that germ cell development in XAP5 cKO males was arrested in meiotic prophase, and XAP5 is indispensable for meiotic progression during spermatogenesis.

Ablation of XAP5 in germ cells results in arrested spermatogenesis and absence of sperm.

(A) Testes sizes, body weight and testis weight from WT and XAP5 cKO mice at various ages were evaluated. Similar body weight was observed between WT and XAP5 cKO mice. No significant difference in size and weight was detected between WT and XAP5 cKO testes until P16. Scale bars: 1 mm. n = 5, ***P < 0.001, n.s. stands for not significant. (B) Histology of testes from WT and XAP5 cKO mice at various ages were evaluated by H&E staining. No pachytene spermatocytes were found, and many vacuolated tubules were observed in XAP5 cKO testis since P16. Black arrowheads indicate pachytene spermatocytes; black arrows indicate round spermatids. Vacuolated seminiferous tubules are indicated by asterisks. Scale bars: 20 μm (black), 2.5 μm (red). (C) Chromosome spreads staining showing pachytene spermatocytes from P16 WT male mouse. Yellow dotted circle indicates the XY body. Scale bars: 10 μm. (D) Immunofluorescence staining of PNA in adult testes indicating the absence of spermatids in XAP5 cKO mice. Scale bars: 20 μm. (E) H&E staining of adult epididymis displaying absence of sperm in XAP5 cKO mice. Scale bars: 50 μm. Error bars depict means ± SEM. All P-values were calculated using an unpaired, two-tailed Student’s t-test.

XAP5 and XAP5L function as antagonistic regulators of ciliary transcriptional regulatory networks via regulating transcription factors

To gain insight into the regulatory mechanisms of XAP5L during sperm flagellar assembly, we employed RNA-seq analysis to examine mature sperm collected from the cauda epididymis of WT and XAP5L KO adult mice. Differential expression analysis identified 2093 upregulated and 267 downregulated genes in XAP5L KO sperm (Figure S3A). Gene ontology analysis revealed that the upregulated genes were associated with cilia assembly or function (Figure 4A and Figures S3B, C). When crossed with the previously compiled ciliogenesis-related gene list(Nemajerova et al., 2016), the upregulated genes were observed encompassing diverse structural and functional components of the ciliogenic program (Table S1 and Figure S3D). Notably, several key transcription factor (TF) regulators of ciliogenesis were among the upregulated genes, including FOXJ1 and RFX families (Figures 4B, C and Figure S3E). Moreover, several key regulators of spermatogenesis, particularly spermiogenesis, were identified, including RFX2(Kistler et al., 2015; Wu et al., 2016), SOX families(Schartl et al., 2018; D. Zhang et al., 2018), TAF7L(Zhou et al., 2013) and TBPL1(Martianov et al., 2001) (Figure 4B and Figure S3E). Many core ciliary genes were also upregulated, such as CFAP206(Beckers et al., 2020) and IFT81(Qu et al., 2020) which are critical for male fertility (Figure S3E), consistent with the tight coregulation between spermiogenesis and cilia-related genes during spermatogenesis. These findings suggest that the genes identified in our study provide an excellent resource for candidates with novel ciliary or spermatogenesis-related functions, and the newly identified testes-specific protein TULP2(Oyama et al., 2022; Zheng et al., 2021) was selected for validation. We observed that TULP2 KO male mice were sterile due to malformation of sperm flagella (Figure S4). Overall, our results clearly indicate that XAP5L acts as a central transcriptional repressor of ciliogenesis during mouse spermatogenesis.

Aberrant ciliogenic and spermatogenic gene expression in XAP5L KO and XAP5 cKO male germ cells.

(A) Gene ontology analysis showing top ten biological process terms significantly enriched among upregulated genes in XAP5L KO sperm. (B) Real-time PCR validation of genes involved in ciliogenesis and spermatogenesis. n = 3, **P < 0.01, ***P < 0.001, n.s. stands for not significant. (C) Detection of FOXJ1 and RFX2 in mature sperm of WT and XAP5L KO mice by Western blot analysis. GAPDH was used as a control. (D) Gene ontology analysis showing top ten biological process terms significantly enriched among downregulated genes in P16 XAP5 cKO testes. (E) Real-time PCR validation of genes involved in ciliogenesis and spermatogenesis. n = 3, *P < 0.05, **P < 0.01, ***P < 0.001, n.s. stands for not significant. (F) Detection of FOXJ1 and RFX2 in P16 testes of WT and XAP5 cKO mice by Western blot analysis. GAPDH was used as a control. (G) XAP5/XAP5L occupancies at FOXJ1 and RFX2 promoters assessed by CUT&Tag-seq (left) and CUT&Tag real-time PCR (middle and right). Red boxes indicate the binding regions of XAP5 and XAP5L. n = 3, *P < 0.05, ***P < 0.001, n.s. stands for not significant. (H) A model on the involvement of XAP5 and XAP5L in the regulation of ciliogenesis during spermatogenesis. Error bars depict means ± SEM. All P-values were calculated using an unpaired, two-tailed Student’s t-test.

Furthermore, we performed RNA-seq analysis using P16 testes samples from WT and XAP5 cKO mice to explore the underlying mechanism of XAP5-mediated germ cell loss. Differential expression analysis identified 554 upregulated and 1587 downregulated genes in XAP5 cKO sperm (Figure S5A). We observed that the downregulated genes in XAP5 cKO males were involved in cilium assembly or functions (Figure 4D and Figures S5B-D). Strikingly, many of these downregulated genes were upregulated in XAP5L KO sperm, including the key ciliogenesis regulators FOXJ1 and RFX families (Figures 4E, F, Figure S5E and Table S1). In addition, meiosis initiation marker STRA8(Anderson et al., 2008; Ferder et al., 2019), meiosis-specific genes SPO11(Romanienko & Camerini- Otero, 2000) and DMC1(Bishop, Park, Xu, & Kleckner, 1992), key transcription regulators of spermiogenesis RFX2(Kistler et al., 2015; Wu et al., 2016), SOX30(D. Zhang et al., 2018) and CREM(Blendy, Kaestner, Weinbauer, Nieschlag, & Schutz, 1996; Nantel et al., 1996) were also downregulated in XAP5 cKO mice (Figures 4E and S5F), consistent with the aforementioned result that loss of XAP5 induced spermatogenesis arrest in meiotic stage.

To identify direct targets of XAP5/XAP5L, we conducted CUT&Tag analyses on P16 mouse testicular germ cells for XAP5, and on P56 mouse testicular germ cells for XAP5L. Our results revealed 23143 peaks enriched for XAP5, compared to only 2290 peaks enriched for XAP5L (Table S1), and the limited number of enriched peaks for XAP5L suggests that the antibody used for XAP5L in the CUT&Tag experiment might have suboptimal performance. A substantial proportion of these peaks were located within promoter regions (Figures S6A, B), indicating that XAP5/XAP5L primarily function to regulate the expression of protein-coding genes. De novo motif analyses showed that the binding motifs for XAP5 and XAP5L shared a conserved sequence (CCCCGCCC/GGGCGGGG) (Figure S6C), suggesting that XAP5 may regulate a subset of the same downstream genes as XAP5L.

To further elucidate the target genes of XAP5 and XAP5L, we integrated our CUT&Tag data with RNA-seq datasets (Table S1). This analysis revealed that among the genes downregulated in XAP5 cKO germ cells, 891 were bound by XAP5 (Figure S6D). Similarly, 75 genes upregulated in XAP5L KO sperm were occupied by XAP5L (Figure S6E). GO analysis indicated that these genes are associated with cilium assembly and related functions (Figures S6D-G). Notably, both FOXJ1 and RFX2 are regulated and occupied by XAP5, with the conserved sequence present within the XAP5-bound regions (Figure 4G). Furthermore, real-time PCR validation of the CUT&Tag experiments confirmed that RFX2 is also occupied by XAP5L, despite the initial CUT&Tag data not showing enriched peaks for the RFX2 gene (Figure 4G). Unfortunately, due to the suboptimal performance of the XAP5L antibody, we did not detect the occupancy of XAP5L on the FOXJ1 promoter. Nevertheless, these findings suggest that XAP5 and XAP5L regulate ciliary gene expression primarily through direct binding to the conserved motif present in these genes.

Considering the evolutionary conservation of XAP5/XAP5L protein across different species and the expression patterns of XAP5, XAP5L, FOXJ1 and RFX factors during spermatogenesis (Figures 1C and S7), these results suggest that XAP5 and XAP5L have antagonistic effects, and they may function upstream of the transcription factor families, including FOXJ1 and RFX factors, to coordinate the ciliogenesis during spermatogenesis (Figure 4H).

Discussion

Cilia and flagella are ancient organelles present in the last eukaryotic common ancestor (LECA)(Mitchell, 2017). Cilia assembly and maintenance are under strict transcriptional regulation in both unicellular and multicellular organisms(Choksi et al., 2014; Collins et al., 2021; Lewis & Stracker, 2021; Thomas et al., 2010). Despite cilia and flagella having an ancient origin, the evolutionary history of ciliary gene regulation has remained an unsolved problem. The main reason is that the master ciliary transcription factors found in multicellular organisms, including FOXJ1 and RFXs, are absent from unicellular organisms(Choksi et al., 2014; Collins et al., 2021; Lewis & Stracker, 2021; Thomas et al., 2010). Previously, we showed that an ancient transcription factor, XAP5, was required for flagella assembly in the unicellular green algae Chlamydomonas(Li et al., 2018). XAP5 proteins are evolutionarily conserved across diverse organisms(Li et al., 2018; Martin-Tryon & Harmer, 2008), which offers the possibility to investigate the conservation of the role of XAP5 in multicellular organisms. In the present study, we report that XAP5 positively regulates transcriptional network of ciliary genes by activating the key regulators including FOXJ1 and RFXs during spermatogenesis.

Cilia are dynamic organelles, and the expression profile of ciliary genes is dynamic during the cell cycle and under certain physiological conditions(Eggenschwiler & Anderson, 2007; Kasahara & Inagaki, 2021; Santos & Reiter, 2008). Cells have developed regulatory mechanisms to generate functional cilia in a temporal and spatial manner(Choksi et al., 2014; Collins et al., 2021; Lewis & Stracker, 2021; Thomas et al., 2010). However, how ciliary transcription factors determine temporal and spatial ciliary transcriptional programs is poorly understood. Sperm flagellar assembly is likewise tightly regulated during a highly complex temporal process named spermatogenesis(Holstein, Schulze, & Davidoff, 2003; Thomas et al., 2010). Several key transcriptional regulators of spermatogenesis, such as FOXJ1 and RFX2, have been identified(Chen, Knowles, Hebert, & Hackett, 1998; Kistler et al., 2015; Wu et al., 2016).

The flagellar formation of FOXJ1-null sperm is severely impaired, and the FOXJ1 target genes are required for sperm motility(Beckers et al., 2020; Chen et al., 1998; Weidemann et al., 2016). RFX2 KO mice are completely sterile and RFX2 regulates the expression of ciliary genes during spermiogenesis(Kistler et al., 2015; Wu et al., 2016). Ciliary transcription factors (FOXJ1 and RFX2) are dynamically regulated during spermatogenesis(Jung et al., 2019). Intriguingly, our results show that XAP5 and XAP5L are two conserved pairs of antagonistic transcription factors that regulate these key transcriptional regulators required for spermatogenesis.

By using Western blot analysis, we found that XAP5 was ubiquitously expressed in all tissues examined, whereas XAP5L expression was restricted to the testes (Figure 1A). Interestingly, previous research found that XAP5L was widely expressed in normal tissue via using TCGA expression data(Thompson et al., 2021). Moreover, loss of XAP5/XAP5L perturbs transcriptional programs in cancer cells. At present, as shown by the recent data, we cannot rule out the possibility that XAP5L is dynamically expressed in other tissues during development. Although XAP5 and XAP5L are highly expressed in testes and critical for spermatogenesis, they are not present in mature sperm (Figure S1). Thus, XAP5/XAP5L are not needed for sperm maintenance. Given the dynamic expression of XAP5/XAP5L during spermatogenesis, it will be interesting to identify the key signaling factors that regulate the timely and spatial expression of XAP5 and XAP5L.

A role for XAP5 in human brain development is suggested by the association of X-linked intellectual disability (XLID) with rare XAP5 missense variants(Lee et al., 2020).

However, how defects in XAP5 lead to XLID syndrome is unknown. Although there is as yet no evidence that XAP5 regulates ciliogenesis in any system besides Chlamydomonas and mice, the function in cilia could help clarify the unexplained role of XAP5 mutations in causing XLID. The possibility that XAP5 could be a ciliary transcription factor in humans would add XLID to the growing list of the second-order ciliopathies.

Materials and Methods

Mice

C57BL/6J strains (from Vital River Laboratories, Beijing, China) and Stra8-GFPCre knockin mice (from Cyagen Biosciences) were used. Floxed-XAP5 mice, XAP5L and TULP2 knockout mice were generated on the C57BL/6J background using the CRISPR/Cas9 system. The sgRNA sequences used to target XAP5, XAP5L and TULP2 were as follows: XAP5L, sgRNA1-GGC TAC CAG AAA CAG GGA CT, sgRNA2-GGG AGT AAG GTC CCC AAA CT; XAP5, sgRNA3-CTA CAG GGC ACT TAT TAA TA, sgRNA4-ATA GTA ATT CCC CCG TGC TT; TULP2, sgRNA5-TGA CTA ATT AGG CCC GAG AG, sgRNA6-GGT TCT TAG AGA GTC AAC GT. Experimental protocols were approved by the ethics committees of Jianghan University. Mice were maintained in a pathogen-free environment with a room temperature of 23 ± 2 °C under a humidity level of 30-70%. The light was maintained on a 12-hour day/night cycle.

Genome PCR

Mouse genomic DNA was isolated from the tail tip using One Step Mouse Genotyping Kit (Vazyme, PD101), and subjected to PCR with 2 × Taq Plus Master Mix (Vazyme, P212) following the manufacturer’s instructions. The PCR products were separated by 2% agarose gel electrophoresis. The primers used are listed in Table S2.

RNA isolation and real-time PCR

Total RNA was isolated using the TRIzol reagent (Invitrogen, 15596018) and then converted to cDNA with HiScript II 1st Strand cDNA Synthesis Kit (Vazyme, R212) following the manufacturer’s instructions. Real-time PCR was conducted using ChamQ SYBR qPCR Master Mix (Vazyme, Q311) on an ABI StepOnePlus real-time PCR system (Applied Biosystems). Expression values were normalized using the ΔΔCt method with Gapdh as the internal control. The primer sequences are indicated in Table S2.

Western blot analysis

Cell lysates from mouse tissues were prepared in RIPA buffer (Beyotime, P0013B) containing Protease Inhibitor Cocktail (Roche, 11873580001) and 1mM PMSF. Subcellular lysates of testicular nuclear and cytoplasmic fractions were prepared using Nuclear and Cytoplasmic Protein Extraction Kit (Beyotime, P0027) following the manufacturer’s instructions. The protein lysates were subjected to standard immunoblotting analysis. Antibodies used were as follows: anti-β-actin (ABclonal, AC026), anti-XAP5 (Prospertech, HU-412003), anti-XAP5 (Sigma, HPA003585), anti- XAP5L (Prospertech, Hu-412001), anti-LMNB (Proteintech, 66095-1-Ig), anti-GAPDH (ABclonal, AC002), anti-FOXJ1 (D360355, Sangon Biotech), anti-RFX2 (K110716P, Solarbio) and anti-TULP2 (Prospertech, Hu-412002).

Immunofluorescence

Mouse testis were fixed in 4% paraformaldehyde in PBS at 4 °C overnight, dehydrated in gradient ethanol, embedded in paraffin, and sectioned into 5 µm slices. After deparaffinization and rehydration, testis sections were boiled in 1 mM EDTA, pH 8.0 for 15 min using a microwave oven for antigen retrieval, followed by cooling to room temperature. The primary antibodies, including anti-XAP5 (Prospertech, HU-412003, 1:250 dilution), anti-XAP5 (Sigma, HPA003585, 1:250 dilution), anti-XAP5L (Prospertech, Hu-412001, 1:250 dilution), anti-lectin PNA, Alexa Fluor 488 Conjugate (ThermoFisher, L21409, 1:250 dilution), and anti-TULP2 (Prospertech, Hu-412002, 1:250 dilution), were then incubated at 4 °C overnight. Subsequently, the sections were incubated with the secondary antibody donkey anti-rabbit Alexa Fluor Plus 488 (Invitrogen, A32790, 1:500 dilution) for 1 hour and DAPI for 5 minutes at room temperature. Images were collected using a Zeiss Axio Vert A1 microscope.

Histological analysis

Testes and epididymides were fixed in Bouin’s solution (sigma, HT10132) overnight at 4 °C, embedded in paraffin and sectioned into 5 µm slices. The sections were then subjected to H&E staining using standard procedures and observed under a Zeiss Axio Vert A1 microscope.

Assessment of sperm counts, motility and morphology

Cauda epididymides were collected from male mice and dissected in 37°C pre-warmed human tubal fluid (HTF) culture medium (EasyCheck, M1130), followed by incubation at 37°C for 15 min to release the sperm. To quantify sperm abundance, the sperm suspension was fixed in 1% paraformaldehyde for 30 min. Sperm were then counted under a Zeiss ×10 bright-field microscope. To assess sperm motility, the sperm suspension was observed and recorded under a Leica total internal reflection fluorescence microscope. Motility was defined as any movement of the sperm flagellum during a 10- second observation cycle. After immobilization with 1% paraformaldehyde, sperm morphology was observed by H&E staining using standard procedures. Images were captured using a Zeiss Axio Vert A1 microscope equipped with a ×100 oil immersion objective. Sperm without staining were also used to assess sperm morphology. For ultrastructural observation, sperm samples were fixed in 3% glutaraldehyde in 0.1M sodium phosphate buffer (pH7.4), and post-fixed with 1% (wt/vol) OsO4. After dehydration, the samples were placed in propylene oxide and embedded in a mixture of Epon 812 and Araldite. Ultrathin sections obtained by a Leica EM UC7 ultramicrotome were stained with uranyl acetate and lead citrate, then analyzed using a HT7700 TEM (Hitachi).

Fertility test

Each of the adult male mice was mated with two females for 2 months. All the mice used were aged 6-8 weeks. Vaginal plugs were checked every morning, and the number of newborn pups was counted.

Chromosome spreads and immunofluorescence

Testicular chromosome spreads were prepared and immunolabeled as described(Reinholdt, Ashley, Schimenti, & Shima, 2004). The primary antibodies used were as follows: anti-γ H2AX (Millipore, 05-636, 1:300 dilution), anti-SYCP3 (Proteintech, 23024-1-AP, 1:300 dilution). Secondary antibodies used were Alexa Dye (AlexaFluor Plus 488/594) conjugates (ThermoFisher) at 1:500 dilutions. Images were collected with a Zeiss Axio Vert A1 microscope.

RNA-seq analysis

The mature sperm samples from P56 mice and testis samples from P16 mice were washed three time in 1× PBS after harvest. Total RNA was extracted from the samples using the TRIzol reagent (Invitrogen) and purified with the NEBNext Poly(A) mRNA Magnetic Isolation Module (NEB, E7490). The quality of the RNA samples was examined using the Agilent 2100 Bioanalyzer system (Agilent Technologies). Qualified RNA samples were subjected to sequencing library construction using the NEBNext Ultra II RNA Library Prep Kit for Illumina (NEB, E7775). Paired-end (2 × 150 bp) sequencing was carried out on the Illumina NovaSeq6000 platform. The sequencing reads were mapped to the Mus musculus reference genome (GRCm38/mm10) using HISAT2 v2.1.0(Kim, Langmead, & Salzberg, 2015), and StringTie v1.3.3b(Pertea et al., 2015) was applied to assemble the mapped reads. Fragments Per Kilobase of transcript per Million fragments mapped (FPKM) was used to measure the expression level of a gene. Differential expression analysis was processed by DESeq2 v1.12.4(Subramanian et al., 2005), and the significant differentially expressed genes were identified with a fold change >2. Gene Ontology analysis was performed using the DAVID database.

CUT&Tag analysis

The testicular germ cells were isolated using a two-step enzymatic digestion process as previously described (Liu et al., 2017). Briefly, after removing the tunica albuginea, the testes were incubated in 1 mg/mL type IV collagenase (Sigma, C5138) at 37℃ for 15 min with gentle shaking every 5 min. The resulting cell suspension was centrifuged at 1000 rpm for 3 min, and the pellet was further digested with 0.25% Trypsin-EDTA (Gibco, 25200072) at 37℃ for 10 min. The digestion was terminated by adding fetal bovine serum (FBS), and the cell suspension was centrifuged at 1000 rpm for 3 min. The cell pellet was resuspended in DMEM medium containing 10% FBS and cultured at 37℃ in a 5% CO2 atmosphere for 3 h. Floating and weakly adherent cells were collected as germ cells. CUT&Tag and CUT&Tag real-time PCR were performed according to the manufacturer’s protocol using the Hyperactive Universal CUT&Tag Assay Kit for Illumina Pro (Vazyme, TD904). Approximately 100,000 germ cells were used per sample. For real-time PCR, IgG sera were used as a negative control for the XAP5 CUT&Tag, while lysates from XAP5L KO testis were incubated with the XAP5L antibody as a negative control for the XAP5L CUT&Tag. DNA spike in was used as the internal control. The primer sequences are indicated in Table S2.

For analysis, quality control and read trimming were performed using FastQC (v0.11.9) with default parameters and Trimmomatic (v0.39) with the following parameters: ILLUMINACLIP:NexteraPE-PE.fa:2:30:10:8:true LEADING:3 TRAILING:3 SLIDINGWINDOW:4:15 MINLEN:8. Clean reads were aligned with bowtie2 (v2.4.4)(Langmead & Salzberg, 2012) against the Mus musculus (mm39) genome with the parameters -X 1000 --very-sensitive. PCR duplicates were removed using picard’s MarkDuplicates with default parameters. Peak calling was conducted using using MACS2 (v2.1.4) with the parameters -f BAMPE -B --SPMR --keep-dup all. Peaks were annotated to the nearest feature using ChIPseeker (v1.32.0)(G. Yu, Wang, & He, 2015). Motif finding and annotation were performed using Homer findMotifsGenome.pl and annotatePeaks.pl functions, respectively(Heinz et al., 2010). Track screen shots were produced in IGV (version 2.18.1)(Robinson, Thorvaldsdottir, Turner, & Mesirov, 2023).

Statistical analysis

Statistical analyses were conducted using GraphPad Prism 8.0.2 software. All data were presented as mean ± SEM. Values of P < 0.05 were considered statistically significant. Statistical significance between two groups was calculated using an unpaired, parametric, two-sided student’s t test.